Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S.

### 2023-04-23 Draft.Rmd
Update Draft.Rmd

  Air pollution has consequences for everyone, most especially for those with pre-existing conditions. Air pollution is measured in tiny particles or droplets in the air that are two and one-half microns or less in width called PM2.5. So being able to predict the average PM2.5 concentrations ($\mu$g/m$^3$) will allow us to keep people most vulnerable safe as we look for and implement changes to improve the air quality in the region. We created four models to predict the annual average of ambient air pollution in a location based on the variables in the dataset. We chose our predictors using the Random Forest algorithm. And for our four models, we chose Linear Regression, K-Nearest Neighbors, Xgboosting, and Random Forest. Linear Regression models a linear relationship between the PM2.5 value in the atmosphere and the predictors. K-Nearest Neighbors returns predicted PM2.5 values based on their neighboring data points. Xgboosting predicts PM2.5 values by combining the estimates of a set of simpler, weaker models. Random forest makes an averaged prediction from a collection of independent decision trees. We hypothesize that the RMSE performance of the Xgboosting model will be the best, with a value of around 0.5.

@@

### Models 
### 1. Linear Regression
For linear regression (LR), the unscaled training set was used since variance does not affect the prediction performance of this model. Step Akaike Information Criteria (AIC) was used to simplify feature amount without impacting the model's performance. Features are dropped as the AIC score decreases and feature dropping stops once the AIC score increases. 'CMAQ'and 'aod' were determined to be the most important features, and these predictors were used in the recipe. Then, model was fitted. Using the summary function, we found the LR model had a RMSE of 2.27 performance. After 10-fold cross validation, the RMSE score was over 1, indicating that this model's prediction performance is not good.

@@

The Extreme Gradient Boosting (XGB) model performs by training weak learners, models that have low prediction accuracy, sequentially to become a single strong learner, a model that has strong prediction accuracy. We preprocess the training set to make the predictive modeling process more accurate and smoother by prepping (estimates the quantities for 'train_scaled') and baking (assigns prepped data to an object) the data. The aforementioned top 10 predictors were used in this preprocessing recipe. Moreover, the categorical variables were factored and features with no variance were removed before this data was baked. Then, this processed data is assigned to 'folds3' and resampled on 5 subsets of the training set. Using 'fold3', we tune the hyperparameters to make the modeling process more efficient and perform grid specification to determine which hyperparameter values have high prediction accuracy (i.e. low prediction error). Next, we isolated the best hyperparameter values, so they can be used for the final boosting model. Lastly, we fitted the model and evaluated its prediction performance. Its RMSE score was 0.69, indicating the XG Boosting is an adequate model. This website was used: [RBloggers](https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/)

@@

The Random Forest (RF) Regression model performs by training multiple different samples simultaneously and averages the outcome at the end for the final prediction. The randomness of these samples contribute to the prediction accuracy of this model. Similar to the XG Boosting Regression model, we preprocess the training set with the same features to improve make the predictive modeling process and then create model specifications. This processed data is assigned to 'folds4' and resampled on 5 subsets of the training set to tune the hyperparameters. We then perform grid specification to determine which hyperparameter values have high prediction accuracy (i.e. low prediction error). Next, we isolated the best hyperparameter values, so they can be used for the final boosting model. Lastly, we fitted the model and evaluated its prediction performance. Its RMSE score was 8.81, indicating the RF is not a good predictive model.

@@

As seen on the table and visual, LR has the highest RMSE and lowest R-squared estimate. KNN has the second highest RMSE with a moderate R-squared estimate. And XGB has the lowest RMSE of 0.69 and the highest R-squared estimate of 0.93. Random Forest model has the both the second lowest RMSE and second highest R-squared estimate. Since we are choosing the best model by the best RMSE, which is the lowest one, the Xgboost model is thus the best and final model based on our procedure. After fitting the Xgboost model to the testing data, we find that its RMSE and R-squared is 1.39 and 0.66, respectively. Since we are trying to predict annual average ambient air pollution of a given monitor, then these statistics mean that the amount of error in our model's predictions is about 1.39 $\mu$g/m$^3$, and that the features we used in our model are able to explain 66% of the variance in the data, which means this model does not fit/predict the data well.

>>>>

Air pollution has consequences for everyone, most especially for those with pre-existing conditions. Air pollution is measured in tiny particles or droplets in the air that are two and one-half microns or less in width called PM2.5. So being able to predict the average PM2.5 concentrations ($\mu$g/m$^3$) will allow us to keep people most vulnerable safe as we look for and implement changes to improve the air quality in the region. We created four models to predict the annual average of ambient air pollution in a location based on the variables in the dataset. We chose our predictors using the Random Forest algorithm. And for our four models, we chose Linear Regression, K-Nearest Neighbors, XG Boost, and Random Forest. Linear Regression models a linear relationship between the PM2.5 value in the atmosphere and the predictors. K-Nearest Neighbors returns predicted PM2.5 values based on their neighboring data points. XG Boost predicts PM2.5 values by combining the estimates of a set of simpler, weaker models. Random forest makes an averaged prediction from a collection of independent decision trees. We hypothesize that the RMSE performance of the XG Boost model will be the best, with a value of around 0.5.

@@

### Models 
### 1. Linear Regression
For linear regression (LR), the unscaled training set was used since variance does not affect the prediction performance of this model. Step Akaike Information Criteria (AIC) was used to simplify feature amount without impacting the model's performance. Features are dropped as the AIC score decreases and feature dropping stops once the AIC score increases. 'CMAQ'and 'aod' were determined to be the most important features, and these predictors were used in the recipe. Then, the model was fitted. Using the summary function, we found the LR model had a RMSE of 2.27 performance. After 10-fold cross validation, the RMSE score was over 1, indicating that this model's prediction performance is not good.

@@

The Extreme Gradient Boosting (XGB) model performs by training weak learners, models that have low prediction accuracy, sequentially to become a single strong learner, a model that has strong prediction accuracy. We preprocess the training set to make the predictive modeling process more accurate and smoother by prepping (estimates the quantities for 'train_scaled') and baking (assigns prepped data to an object) the data. The aforementioned top 5 predictors were used in this preprocessing recipe. Moreover, the categorical variables were factored and features with no variance were removed before this data was baked. Then, this processed data is assigned to 'folds3' and resampled on 5 subsets of the training set. Using 'fold3', we tune the hyperparameters to make the modeling process more efficient and perform grid specification to determine which hyperparameter values have high prediction accuracy (i.e. low prediction error). Next, we isolated the best hyperparameter values, so they can be used for the final boosting model. Lastly, we fitted the model and evaluated its prediction performance. Its RMSE score was 0.69, indicating that XG Boost is an adequate model. This website was used: [RBloggers](https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/)

@@

The Random Forest (RF) Regression model performs by training multiple different samples simultaneously and averages the outcome at the end for the final prediction. The randomness of these samples contribute to the prediction accuracy of this model. Similar to the XG Boosting Regression model, we preprocess the training set with the same features to improve the predictive modeling process and then create model specifications. This processed data is assigned to 'folds4' and resampled on 5 subsets of the training set to tune the hyperparameters. We then performed grid specification to determine which hyperparameter values have high prediction accuracy (i.e. low prediction error). Next, we isolated the best hyperparameter values, so they can be used for the final boosting model. Lastly, we fitted the model and evaluated its prediction performance. Its RMSE score was 8.81, indicating the RF is not a good predictive model.

@@

As seen on the table and visual, Random Forest has the highest RMSE and lowest R-squared estimate. LR has the second highest RMSE with a moderate R-squared estimate. And XG Boost has the lowest RMSE of 0.23 and the highest R-squared estimate of 0.99. The kNN model has the both the second lowest RMSE and second highest R-squared estimate. Since we are choosing the best model by the best RMSE, which is the lowest one, the XG Boost model is thus the best and final model based on our procedure. After fitting the XG Boost model to the testing data, we find that its RMSE and R-squared is 1.38 and 0.68, respectively. Since we are trying to predict annual average ambient air pollution of a given monitor, then these statistics mean that the amount of error in our model's predictions is about 1.38 $\mu$g/m$^3$, and that the features we used in our model are able to explain 68% of the variance in the data, which means this model does not fit/predict the data well.
